---
title: "Reproducible Research: Peer Assessment 1"
author: Patricia J. Hawkins
output: 
  html_document:
  keep_md: true
---
```{r global_options, include=FALSE}
library(knitr)
#opts_chunk$set(fig.width=12, fig.height=8, fig.path='Figs/', warning=FALSE, message=FALSE) # don't include warnings and errors in output document
opts_chunk$set(fig.width=10, fig.height=8)
par(bg = "white")
```

Current timestamp: `r date()` 

## Loading and preprocessing the data

```{r read_file}
library(data.table)
library(lubridate)

activities <- as.data.table(read.csv("./activity.csv"))
activities$date <- ymd(activities$date)
```

## What is mean total number of steps taken per day?
```{r mean}
library(plyr)
stats <- ddply(activities, ~date,summarise,total_steps_per_day = sum(steps,na.rm=TRUE),
               mean=mean(steps, na.rm = TRUE),
               median=median(steps, na.rm = TRUE))
```
###  1. Make a histogram of the total number of steps taken each day
```{r hist}
hist(stats$total_steps_per_day, breaks = 15, bg="white")
```
### 2. Calculate and report the mean and median total number of steps taken per day

#### Mean number of steps taken per day: `r mean(stats$total)`

#### Median number of steps taken per day: `r median(stats$total)`

## What is the average daily activity pattern?

### 1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, across all days (y-axis)
```{r plot average steps per day}
average_daily <- as.data.table(aggregate(steps ~ interval, activities, "mean"))
plot(average_daily, type = "l", bg="white")

```
### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

#### Interval `r average_daily[steps == max(average_daily$steps)]$interval` contains `r max(average_daily$steps)` steps on average

## Imputing missing values

### 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
####  Number of rows with missing values: `r sum(!complete.cases(activities))`
### 2. For missing values, use the mean value for that interval, which we calculated for the average daily activity pattern.
### 3. Create new dataset new_activities; fill in missing step values.

``` {r fill_missing_values}
setkey(average_daily, interval)

new_activities <- activities
setkey(new_activities, date, interval)
cases <- new_activities[!complete.cases(activities)]

for (i in 1:nrow(cases)) {              # Should be possible to vectorize for-loop,  since none of the results depend on each other, but this works.
    row <- cases[i]
    steps <- as.integer(average_daily[.(row$interval)]$steps + .5) # round to nearest integer
    new_activities[.(row$date, row$interval)]$steps <-  steps
 }
```
### 4. Make a histogram of the total number of steps taken each day

``` {r hist_total_daily_steps }
total_daily <- as.data.table(aggregate(steps ~ date, new_activities, "sum"))
hist(total_daily$steps, breaks = 15)
```
#### 4a. Calculate and report the mean and median total number of steps taken per day.
#### Mean total steps per day:  `r mean(total_daily$steps)`
#### Median total steps per day: `r median(total_daily$steps)`

## Are there differences in activity patterns between weekdays and weekends?
### 1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
``` {r day of week factor}

new_activities$day_of_week <- factor(as.integer(wday(new_activities$date) %in% list(1,7)),
                                     levels = c(0,1), labels=c("weekday", "weekend"))
```
### Make a panel plot containing a time series plot (i.e. type = "l") of
### * the 5-minute interval (x-axis)
### * and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r panel plot}

library(ggplot2)
qplot(interval, steps, data = aggregate(steps ~ interval + day_of_week, new_activities, "mean"), facets = ~day_of_week) + geom_line()

```










